Hydra battles in Coq

### 2020-09-28 doc/hydras.tex
correcting typos

\documentclass[a4paper]{book}
\usepackage{alltt}
\usepackage{makeidx}
\usepackage{hyperref}
\usepackage{url}
\usepackage{mdframed}
\usepackage{amsmath}
\usepackage{varioref}
\usepackage{synttree, proof, centernot}
\usepackage[firstpage]{draftwatermark}
\usepackage{verbatim}
\usepackage[note]{marginote}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{dsfont}
\usepackage{threeparttable}
\setcounter{tocdepth}{2}
\input{coqmacros}

@@

Proof assistants are excellent tools for exploring the structure of mathematical proofs,
studying  which hypotheses are really needed, and which proof patterns are useful and/or
necessary. Since the development of a theory is represented as a bunch of computer files,
everyone is able to read the proofs with an arbitrary level of detail, or to play with the theory by building alternate proofs or definitions.

@@

\item We prove a relation between the length of a ``classic''  class of  battles
and the Wainer-Hardy hierarchy of ``rapidly growing functions'' $H_\alpha$~\cite{Wainer1970}. The considered class of battles, which we call \emph{standard}  is the most considered one in the scientific  litterature(including popularization).
\end{itemize}

@@

It may often happen that several definitions of a given concept, or several proofs of a given theorem are possible. If all the versions present some interest, we will make them available, since each one may be of some methodological 
interest (by illustrating some tactic of proof pattern, for instance).
We may use \coq's module system to make several proofs of a givent theorem co-exists in our libraries (see also Sect~\vref{sect:alt-proofs}).
After some discussions of the pros and cons of each solution, we will choose to develop only one of them, leaving the others  as exercises or projects (i.e., big or difficult exercises).
In order to discuss which assumptions are really needed for proving a theorem, we will also present 
several aborted proofs.
Of course, do not hesitate to contribute nice proofs or alternative definitions !

@@

Obviously, the third theorem implies the second one, which implies the first one. So, theoretically, a library would contain only a proof of $(3)$ and remarks for $(2)$ and $(1)$. But we found it interesting to make all the three proofs available, allowing the reader to compare their common structure and notice their technical differences.
In particular, the proof of $(3)$ uses several non-trivial combinatorial properties of ordinal numbers~\cite{KS81}, whilst $(1)$ and $(2)$ use simple properties of $\mathbb{N}$ and $\mathbb{N}^2$.

@@

At any place of our development, you may use the command \texttt{Print Assumptions {\it ident}} in order to verify on which axiom the theorem {\it ident} may depend. The following example is extracted from 
Library~\href{../src/html/hydras.Epsilon0.F_alpha.html}{hydras.Epsilon0.F\_alpha}, where we use the \texttt{coq-equations} plug-in (see Sect.~\vref{sect:wainer}).

@@

\subsection{Contributions}
Many thanks to \'Evelyne Contejean and Florian Hatat for their contribution. \'Evelyne contributed libraries on the recursive path ordering (\emph{rpo}) for proving the well-foundedness of our representation of $\epsilon_0$ and $\Gamma_0$.
Florian Hatat proved many useful lemmas on countable sets, which we used in our adaptation of Schütte's formalization of countable ordinals.

@@

Any form of contribution  is welcome: correction of errors, improvement of
Coq scripts, proposition of inclusion of new chapters, and generally any
comment or proposition that would help us. The text contains several \emph{projects} which, when completed, may improve the present work.
Please do not hesitate to bring your contribution!

@@

\subsection{Alternative or bad definitions}
\label{sect:alt-proofs}
Finally, we decided to include definitions or lemma statements, as well as tactics,  that lead to
dead-ends or to too complex developments, with the following coloring.
Bad definitions 
 and encapsulation in modules called \texttt{Bad}, \texttt{Bad1}, etc.

@@

\label{sec:orgheadline44}
A \emph{hydra battle} is a fight between Hercules and the Hydra. 
More formally, a  battle is a sequence of \emph{rounds}.
At each round:
\begin{itemize}
\item If the hydra is composed of just one head, the battle is finished
and  Hercules is the winner.
\item Otherwise, Hercules chops off \emph{one} head of the hydra,

\begin{itemize}
\item If the head is at distance 1 from the foot, the head is just lost by the hydra, with no more reaction.
\item Otherwise, let us denote by \(r\) the node that was at distance \(2\) from 
the removed head in the direction of the foot,  and consider the  sub-hydra \(h'\) of \(h\), whose  root is \(r\) \footnote{$h'$ will be called ``the wounded part of the hydra'' in the subsequent text. In Figures~\vref{fig:Hy2} and ~\vref{fig:Hy4}, this sub-hydra  is displayed in red.}. Let $n$ be some natural number.
Then $h'$ is replaced by  $n+1$ of copies of \(h'\) which share the same root $r$.
 The \emph{replication number} $n$ may be different (and generally is)   at each round of the fight.
It may be chosen by the hydra, according to its strategy, or imposed by some 
particular rule. In many presentations of hydra battles, this number is increased by $1$ at each round. In the following presentation, we will also consider battles where the hydra is free to chose its number of replication at every round of the battle\footnote{Let us recall that, if the chopped-off head was at distance 1 from the foot, the replication factor is meaningless.}.
\end{itemize}
\end{itemize}

@@

This goal is easily solvable, using some arithmetic. We let the reader look at the source of this development.

@@

The mutually exclusive cases we consider are the following.
\begin{itemize}
\item \textbf{R1}: The chopped off head was at distance 1 from the foot.
\item \textbf{R2}: The chopped off head was at a distance greater than or equal to  $2$ from the foot.
\end{itemize}

@@

If Hercules chops off a head near the floor, there is no replication at all. We use an auxiliary predicate, associated with the removing of one head from a sequence of hydras.

@@

Let us now consider beheadings  where the chopped-off head is at distance greater than or equal to $2$ from the foot. All the following relations are parameterized by the number $n$ of new copies added by the hydra.

@@

\subsection{Description of any round}

@@

\textbf{Hint:} Please keep in mind that the last  configuration of your interactively built battle is known only at the end of the battle. Thus, you will have to create and solve subgoals with existential variables. For that purpose, the tactic \texttt{eexists}, applied to the 
goal \Verb@{h':Hydra | h -*-> h'}@ generates the subgoal \Verb|h -*-> ?h'|.
\end{exercise}

@@

\begin{itemize}
\item Hercules chops always the rightmost head of the hydra.
\item The battle is standard: at the round number $i$, the expected replication is $i$.
\end{itemize}

@@

The following lemma  is a formal statement about these first rounds, in terms of the
\texttt{battle} predicate.

@@

We can make bigger steps through iterations of \texttt{next}.

The functional \texttt{iterate}, similar to Standard Library's \texttt{Nat.iter},
is defined and studied in~\href{../src/html/hydras.Prelude.Iterates.html\#iterate}{Prelude.Iterates}).

@@

The battle we study seems to be awfully long. Let us concentrate our
tests on some particular events : the states where $\texttt{nh}=0$

@@

We now have some intuition of the sequence.
It looks like the next ``\texttt{nh}=0'' event will happen at the $192=2(95+1)$-th round, then at the $2(192+1)$-th round.

@@

\begin{Coqsrc}
Inductive one_step (i: nat) :
  nat -> nat -> nat -> nat -> nat -> nat -> Prop :=
| step1 : forall a b c,  one_step i a b (S c) a b c
| step2 : forall a b ,  one_step i a (S b) 0 a b (S i)
| step3  : forall a, one_step i (S a) 0 0 a (S i) 0.
\end{Coqsrc}

@@

\subsection{Two small proofs of impossibility}
\index{Maths!Proofs of impossibility}

@@

\subsubsection{No variant on Peano.lt}
\label{omega-case}

@@

Unfortunately, we can prove that 
\emph{no} instance of class \texttt{WfVariant round Peano.lt $m$} can be built, where
$m$ is \emph{any} function of type \texttt{Hydra $\arrow$ nat}.

@@

Let us assume there is a variant $m$ from \texttt{Hydra} into \texttt{nat} for proving
    the  termination of all hydra battles.

@@

 Using the functions $m$ and $\iota$, we define a second hydra \texttt{small\_h}, and show
 there is a one-round battle that transforms \texttt{big\_h} into \texttt{small\_h}. Please note that,
due to the hypothesis \texttt{Hvar}, we are interested in the termination of \emph{free} battles. 
There is no problem to consider a round with \texttt{m big\_h} as replication factor.

@@

\begin{Coqsrc}
Lemma m_ge : forall i:nat , i <= m (iota i).
\end{Coqsrc}

@@

\subsection{A proof a termination using \texorpdfstring{$\omega^2$}{omega^2}} 
\label{sect:merge-example}.

@@

\index{Projects}
\begin{project}
\label{project:succ-limit-dec}
Add to the class \texttt{ON} the requirement that for any $\alpha$ it is decidable whether $\alpha$ is $0$, a successor or a limit ordinal.
\end{project}

@@

\section{Accessibility inside \texorpdfstring{$\epsilon_0$}{epsilon0} : Paths}
\index{Maths!Accessibility inside epsilon0}
\label{sect:pathes-intro}

@@

\section{Large  sequences}

\subsection{Definition}

@@

\section{Computing mlarge sequences}

@@

\subsection{Looking for a function definition}

@@

\subsection{Properties derived from \texttt{L\_spec}}

@@

\section{Using \texttt{Equations}}
\label{sect:L-equations}

@@

\section{The Wainer-Hardy hierarchy}
\label{sect:hardy}
In order to give an idea of the complexity of the functions  $L_\alpha$s, we compare them with a better known family of functions, the so called \emph{Wainer-Hardy hierarchy} of fast growing functions,
presented for instance in~\cite{Promel2013}. 
\index{Maths!Hardy Hierarchy}

@@

\subsubsection{Project \vref{project:succ-limit-dec}}
\paragraph*{Hint:} 
  Beware of the instances associated with sum and product of notations !
  You may consider additional fields such as ``there is a greatest element'' 
to make the sum and product of notations ``compositional''.

>>>>

\documentclass[a4paper]{book}
\usepackage{alltt}
\usepackage{makeidx}
\usepackage{hyperref}
\usepackage{url}
\usepackage{mdframed}
\usepackage{amsmath}
\usepackage{varioref}
\usepackage{synttree, proof, centernot}
\usepackage[firstpage]{draftwatermark}
\usepackage{verbatim}
\usepackage[note]{marginote}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{dsfont}
\usepackage{threeparttable}
\setcounter{tocdepth}{1}
\input{coqmacros}

@@

Proof assistants are excellent tools for exploring the structure of mathematical proofs,
studying  which hypotheses are really needed, and which proof patterns are useful and/or
necessary. Since the development of a theory is represented as a bunch of computer files,
everyone is able to read the proofs with an arbitrary level of detail, or to play with the theory by writing alternate proofs or definitions.

@@

\item We prove a relation between the length of a ``classic''  class of  battles \footnote{This class is also called \emph{standard} in this document (text and proofs). The \emph{replication factor} of the hydra is exactly $i$ at the $i$-th round of the battle (see Sect~\vref{sect:replication-def}).}
and the Wainer-Hardy hierarchy of ``rapidly growing functions'' $H_\alpha$~\cite{Wainer1970}. The considered class of battles, which we call \emph{standard}  is the most considered one in the scientific  litterature(including popularization).
\end{itemize}

@@

It may often happen that several definitions of a given concept, or several proofs of a given theorem are possible. If all the versions present some interest, we will make them available, since each one may be of some methodological 
interest (by illustrating some tactic of proof pattern, for instance).
We may use \coq's module system to make several proofs of a given theorem co-exist in our libraries (see also Sect~\vref{sect:alt-proofs}).
After some discussions of the pros and cons of each solution, we develop only one of them, leaving the others  as exercises or projects (i.e., big or difficult exercises).
In order to discuss which assumptions are really needed for proving a theorem, we will also present 
several aborted proofs.
Of course, do not hesitate to contribute nice proofs or alternative definitions !

@@

Obviously, the third theorem implies the second one, which implies the first one. So, theoretically, a library would contain only a proof of $(3)$ and remarks for $(2)$ and $(1)$. But we found it interesting to make all the three proofs available, allowing the reader to compare their common structure and notice their technical differences.
In particular, the proof of $(3)$ uses several non-trivial combinatorial properties of ordinal numbersup to $\epsilon_0$~\cite{KS81}, whilst $(1)$ and $(2)$ use simple properties of $\mathbb{N}$ and $\mathbb{N}^2$.

@@

At any place of our development, you may use the  \texttt{Print Assumptions {\it ident}} command in order to verify on which axiom the theorem {\it ident} may depend. The following example is extracted from 
Library~\href{../src/html/hydras.Epsilon0.F_alpha.html}{hydras.Epsilon0.F\_alpha}, where we use the \texttt{coq-equations} plug-in (see Sect.~\vref{sect:wainer}).

@@

\subsection{Contributions}
Many thanks to \'Evelyne Contejean,  Florian Hatat and Pascal Manoury  for their contribution. \'Evelyne contributed libraries on the recursive path ordering (\emph{rpo}) for proving the well-foundedness of our representation of $\epsilon_0$ and $\Gamma_0$.
Florian Hatat proved many useful lemmas on countable sets, which we used in our adaptation of Schütte's formalization of countable ordinals. Pascal Manoury is integrating the ordinal $\omega^\omega$ into our hierarchy of ordinal notations.

@@

Any form of contribution  is welcome: correction of errors (typos and more serious mistakes), improvement of
Coq scripts, proposition of inclusion of new chapters, and generally any
comment or proposition that would help us. The text contains several \emph{projects} which, when completed, may improve the present work.
Please do not hesitate to bring your contribution!

@@

\subsection{Alternative or bad definitions}
\label{sect:alt-proofs}
Finally, we decided to include definitions or lemma statements, as well as tactics,  that lead to
dead-ends or too complex developments, with the following coloring.
Bad definitions 
 and encapsulation in modules called \texttt{Bad}, \texttt{Bad1}, etc.

@@

\label{sec:orgheadline44}
\label{sect:replication-def}

A \emph{hydra battle} is a fight between Hercules and the Hydra. 
More formally, a  battle is a sequence of \emph{rounds}.
At each round:
\begin{itemize}
\item If the hydra is composed of just one head, the battle is finished
and  Hercules is the winner.
\item Otherwise, Hercules chops off \emph{one} head of the hydra,

\begin{itemize}
\item If the head is at distance 1 from the foot, the head is just lost by the hydra, with no more reaction.
\item Otherwise, let us denote by \(r\) the node that was at distance \(2\) from 
the removed head in the direction of the foot,  and consider the  sub-hydra \(h'\) of \(h\), whose  root is \(r\) \footnote{$h'$ will be called ``the wounded part of the hydra'' in the subsequent text. In Figures~\vref{fig:Hy2} and ~\vref{fig:Hy4}, this sub-hydra  is displayed in red.}. Let $n$ be some natural number.
Then $h'$ is replaced by  $n+1$ of copies of \(h'\) which share the same root $r$.
 The \emph{replication factor} $n$ may be different (and generally is)   at each round of the fight.
It may be chosen by the hydra, according to its strategy, or imposed by some 
particular rule. In many presentations of hydra battles, this number is increased by $1$ at each round. In the following presentation, we will also consider battles where the hydra is free to choose its ~replication factor at each round of the battle\footnote{Let us recall that, if the chopped-off head was at distance 1 from the foot, the replication factor is meaningless.}.
\end{itemize}
\end{itemize}

@@

This goal is easily solvable, using some arithmetic. We let the reader look at the source.

@@

The mutually exclusive cases we consider are the following:
\begin{itemize}
\item \textbf{R1}: The chopped off head was at distance 1 from the foot.
\item \textbf{R2}: The chopped off head was at a distance greater than or equal to  $2$ from the foot.
\end{itemize}

@@

If Hercules chops off a head near the floor, there is no replication at all. We use an auxiliary 
predicate \texttt{S0}, associated with the removing of one head from a sequence of hydras.

@@

Let us now consider beheadings  where the chopped-off head is at distance greater than or equal to $2$ from the foot. All the following relations are parameterized by the replication factor  $n$.

@@

\subsection{Relation associated with a round}

@@

\textbf{Hints:} 
\begin{itemize}

\item Please keep in mind that the last  configuration of your interactively built battle is known only at the end of the battle. Thus, you will have to create and solve subgoals with existential variables. For that purpose, the tactic \texttt{eexists}, applied to the 
goal \Verb@{h':Hydra | h -*-> h'}@ generates the subgoal \Verb|h -*-> ?h'|.
\item You may use Gérard Huet's \emph{zipper} data structure~\cite{zipper} for writing tactics associated with Hercule's  interactive search of a head to chop off.
\end{itemize}

@@

\begin{itemize}
\item Hercules chops off always the rightmost head of the hydra.
\item The battle is standard: at the round number $i$, the expected replication is $i$.
\end{itemize}

@@

The following lemma  is a formal description of these first rounds, in terms of the
\texttt{battle} predicate.

@@

We can make bigger steps through iterations of \texttt{next}.
The functional \texttt{iterate}, similar to Standard Library's \texttt{Nat.iter},
is defined and studied in~\href{../src/html/hydras.Prelude.Iterates.html\#iterate}{Prelude.Iterates}.

@@

The battle we are studying seems to be awfully long. Let us concentrate our
tests on some particular events : the states where $\texttt{nh}=0$

@@

We now have some intuition of the sequence.
It looks like the next ``\texttt{nh}=0'' event will happen at the $192=2(95+1)$-th round, then at the $2(192+1)$-th round, etc.

@@

\begin{Coqsrc}
Inductive one_step (i: nat) :
  nat -> nat -> nat -> nat -> nat -> nat -> Prop :=
| step1: forall a b c, one_step i a b (S c) a b c
| step2:  forall a b, one_step i a (S b) 0 a b (S i)
| step3: forall a, one_step i (S a) 0 0 a (S i) 0.
\end{Coqsrc}

@@

\subsection{A  small proof of impossibility}
\index{Maths!Proofs of impossibility}

\label{omega-case}

@@

@@

Unfortunately, we can prove that 
\emph{no} instance of class (\texttt{WfVariant round Peano.lt $m$}) can be built, where
$m$ is \emph{any} function of type \texttt{Hydra $\arrow$ nat}.

@@

Let us assume there exists some variant $m$ from \texttt{Hydra} into \texttt{nat} for proving
    the  termination of all hydra battles.

@@

 Using the functions $m$ and $\iota$, we define a second hydra \texttt{small\_h}, and show
 there is a one-round battle that transforms \texttt{big\_h} into \texttt{small\_h}. Please note that,
due to the hypothesis \texttt{Hvar}, we are interested in the termination of \emph{free} battles. 
There is no problem to consider a round with (\texttt{m big\_h}) as the replication factor.

@@

\begin{Coqsrc}
Lemma m_ge : forall i:nat, i <= m (iota i).
\end{Coqsrc}

@@

\subsection{A proof of termination using \texorpdfstring{$\omega^2$}{omega^2}} 
\label{sect:merge-example}.

@@

\index{Projects}
\begin{project}
\label{project:succ-limit-dec}
Add to the class \texttt{ON} the requirement that for any $\alpha$ it is decidable whether $\alpha$ is $0$, a successor or a limit ordinal.

\textbf{Hint:}   Beware of the instances associated with sum and product of notations !
  You may consider additional fields such as ``there is a greatest element'' 
to make the sum and product of notations ``compositional''.

@@

\section{Accessibility inside \texorpdfstring{$\epsilon_0$}{epsilon0} : paths}
\index{Maths!Accessibility inside epsilon0}
\label{sect:pathes-intro}

@@

\section{Definitions}

%\subsection{Definition}

@@

\section{The length of minimal large sequences}

@@

%\subsection{Looking for a function definition}

@@

\subsection{Abstract properties}

@@

\subsection{Using \texttt{Equations}}
\label{sect:L-equations}

@@

\section{The Wainer-Hardy hierarchy (functions \texorpdfstring{$H_\alpha$}{\texttt{H\_alpha}})}

\label{sect:hardy}
In order to give an idea of the complexity of the functions  $L_\alpha$s, we compare them with a better known family of functions, the so called \emph{Wainer-Hardy hierarchy} of fast growing functions,
presented for instance in~\cite{Promel2013}. 
\index{Maths!Hardy Hierarchy}

@@
